# Basic R Programming (Part I)  {#programming1}

Today we are going to learn some basic programming concepts. By the end of the session, you should be able to:

-   Run and troubleshoot commands in the R console.

-   Understand the different data types in R and how to create them.

-   Write comments to your code to better structure it.

-   Create and change variables.

-   Create and use functions.

-   Install R packages.

-   To use the help function in R.

## Using the R Console

In the previous chapter, I made a distinction between the script and the console. I said that the script was an environment where we would write and run polished code, and the R console is an environment for writing and running "dirty" quick code to test ideas, or code that we would run one.

That distinction is kinda true, but it's not completely true. In reality, when we write a script we are writing ***commands*** for R to ***execute*** in the R console. When we "run" a script, we are feeding that script to the console. In this sense, the R script is equivalent to a waiter. We tell the waiter (R script) what we want to order, and then the waiter hands that order to the chef (R console).

So even though we will primarily not use the R console in this class, it's important to know how to work with it. We do not want the chef spitting in our food.

### Typing Commands in the Console

The R console uses the operator "\>" to indicate that it is ready for a new command. We enter in our code after this operator and press enter to compute it.

We can command the R console to compute calculations.

```{r eval=FALSE}

> 10 + 20

[1] 30

```

```{r eval=FALSE}

> 20 / 10

[1] 2

```

If you are performing calculations in R, it's important to know that it follows the usual arithmetic convention of order of operations (remember [BIMDAS - Bracets, Indices, Multiplication, Division, Addition, and Subtraction?](https://www.tes.com/en-ie/teaching-resource/bidmas-bodmas-bedmas-bimdas-pemdas-permdas-11154272#:~:text=%E2%80%A2%20BIMDAS%20%2D%20Brackets%2C%20Indices%2C,Multiplication%2C%20Division%2C%20Addition%2C%20Subtraction)).

```{r eval = FALSE}


> (20 + 10 / 10) * 4 

[1] 84

> ((20 + 10) / 10) * 4

[1] 12

```

Now you'll have noticed that the output of every line of code we entered starts with a [1] before our actual result. What does this mean?

Think of the square brackets with a number as a way for R to label and organize its responses. Imagine you have a conversation with R, and every time you ask it something, it gives you an answer. The square brackets with a number, like **`[1]`**, are like labels on each response, telling you which answer corresponds to which question. This is R ***indexing*** its answer.

In each of the above examples, we asked R one question ***per each*** command, which is why the answer is always [1]. If we entered longer code with multiple questions, then we could multiple answers. We could tell which answer related to which question through the index. This is really useful when we ask R long and more complicated questions.
